---
title: 'TRS-BIOME: Forest plots from OSCA'
author: "Svetlina Vasileva"
date: "`r Sys.Date()`"
output:
  epuRate::UQ:
    toc: yes
    number_sections: yes
linkedin: Svetlina Vasileva
twitter: sunsvet23
mail: svetlinav@gmail.com
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(dev = "pdf", #will save graphs in pdf
                      dpi = 300,
                      warning = FALSE,
                      echo=FALSE,
                      message = TRUE, 
                      cache = TRUE,
                      fig.show='hold') #saves output
```

# Packages and files

```{r packages and cleaning,message = FALSE}
library(gridExtra)
library(ggplot2)
library(dplyr)
library(scales)
library(DT)
library(writexl)

data_dir <- "C:/Users/uqsvasil/Documents/PhD/MIA/OSCA/mia_rsq_stats_main.csv"
secondary_dir<-"C:/Users/uqsvasil/Documents/PhD/MIA/OSCA/mia_rsq_stats_secondary.csv"

```

# Wrangling

## Main

```{r}
df <- read.csv(data_dir, header = T, as.is = T)
df$level<-gsub("_clrcheck","",df$level)# remove "clrcheck from the name
df <- df  %>% 
  dplyr::select(type,level,sample,pheno,covs,X..V.O..Vp....Variance..,X..V.O..Vp....SE..,Pval, n) %>%
  dplyr::rename("trait"="pheno","variance"="X..V.O..Vp....Variance..","se"="X..V.O..Vp....SE..", "p_value"= "Pval") %>%
  mutate(covs = gsub("sex_age_dietPC", "age_sex_dietPC", covs)) %>% # change "sex_age_dietPC" to "age_sex_dietPC" 
  mutate(pe=variance*100, 
         pe_se=se*100,
         covs_dummy=case_when(covs=="no_covs" ~ "no_covs",
                              covs!="no_covs" ~  "yes_covs"), #create a dummy covs variable where different types of covs are grouped together
         covs_dummy = factor(covs_dummy, levels = c("yes_covs", "no_covs")))

# Colour per level if significant
df$sig_color <- ifelse(df$p_value < 0.05, as.character(df$level), "black") # in order to color depending on both level and significance

# Only clr-transformed variables
data<-df %>%
  filter(!grepl('_01', level),  # exclude the variables which are not clr-transformed 
         level %in% c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common", "Microba_common"), 
         sample != "no_asd", # for them n=40, it's actually just the unrelated, sib, not powered -> remove
         sample != "full_bristol") %>%   # the full_sample analyses with dietPCs as covs is already that
  mutate(level = factor(level, levels = c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common", "Microba_common")))

```

## Secondary - 08052024

```{r}

df <- read.csv(secondary_dir, header = T, as.is = T)
df$level<-gsub("_clrcheck","",df$level)# remove "clrcheck from the name
df <- df  %>% 
  dplyr::select(type,level,sample,pheno,covs,X..V.O..Vp....Variance..,X..V.O..Vp....SE..,Pval, n) %>%
  dplyr::rename("trait"="pheno","variance"="X..V.O..Vp....Variance..","se"="X..V.O..Vp....SE..", "p_value"= "Pval") %>%
  mutate(covs = gsub("sex_age_dietPC", "age_sex_dietPC", covs)) %>% # change "sex_age_dietPC" to "age_sex_dietPC" 
  mutate(pe=variance*100, 
         pe_se=se*100,
         covs_dummy=case_when(covs=="no_covs" ~ "no_covs",
                              covs!="no_covs" ~  "yes_covs"), #create a dummy covs variable where different types of covs are grouped together
         covs_dummy = factor(covs_dummy, levels = c("yes_covs", "no_covs")))

# Colour per level if significant
df$sig_color <- ifelse(df$p_value < 0.05, as.character(df$level), "black") # in order to color depending on both level and significance

# Only clr-transformed variables
data2<-df %>%
  filter(!grepl('_01', level),  # exclude the variables which are not clr-transformed 
         level %in% c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common", "Microba_common"), 
         covs %in% c("no_covs", "age_sex_dietPC")) %>%   # the full_sample analyses with dietPCs as covs is already that
  mutate(level = factor(level, levels = c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common", "Microba_common")))

```

# Setting up 

## Colour pallete

```{r}
# Generate only for the ones needed
color_9 <- c("#F8766D", "#A3A500", "#39B600", "#00BF7D", "#00BFC4", "#00B0F6", "#9590FF", "#E76BF3", "#FF62BC")

```

## Labels
```{r}

new_labels_levels <-c(
  "species_common" = "Species_common", 
  "species_rm0_rare" = "Species_rare", 
  "genus_common" = "Genus_common",
  "family_common" = "Family_common", 
  "EClevel3_common" = "Enzymes(ECL3)_common", 
  "EClevel4_common" = "Enzymes(ECL4)_common", 
  "MetaCycpathway_common" = "Pathways(MetaCyc)_common", 
  "TCDB_common" = "Transporters(TCDB)_common",
  "Microba_common" = "Genes(Microba)_common"
  )

new_labels_traits <- c(
    "mia_indication" = "MIA indication",
    "mats_indication" = "MatS indication",
    "mia_number" = "MIA number",
    "mats_number" = "MatS number",
    "me_indication" = "ME indication",
    "me_number" = "ME number",
    "ASD" = "ASD", 
    "participant_type" = "Participant type",
    "sleep_cshq_total" = "Sleep",
    "devq_wisc_msel" = "devq_wisc_msel", 
    "age" = "Age",
    "sex" = "Sex",
    "bmi" = "BMI",
    "bristol_regroup" = "Stool type",
    "dietPC1" = "Dietary PC1",
    "dietPC2" = "Dietary PC2",
    "dietPC3" = "Dietary PC3",
    "meds_antibiotic_current" = "Antibiotics current",
    "meds_antipsychotic" = "Antipsychotics",
    "meds_gastrointestinal" = "GI meds",
    "cam_probiotic" = "Probiotic"
    )

new_labels_samples <-c(
  "full_sample" = "Full", # 174
  "174_sample" = "Full ", # 174
  "full_bristol" = "With bristol", #(N=165) 
  "no_sib" = "No SIB", # (N=132)
  "no_asd" = "UNR", # (N=40)
  "no_current_ab" = "No AB", # N=164
  "no_probiotic" = "No PB",  #N=133
  "no_aap" = "No AAP", # (N=169)
  "no_gastro_meds" = "No GI meds", # 156
   "early_age_group" = "Early", # N=64
  "middle_age_group" = "Intermediate",  # N=68
  "late_age_group" = "Older",  # N=42
  "mats_excluded" = "True negative",  # N=90
  "mia_excluded" = "True negative", #N=81
  "acute_negative" = "Acute MIA" # N=83
  )

```

## Plot functions

```{r}
# Main plots

make_plot_covs<-function(dataset, color_column){
  ggplot(dataset, aes_string(y = "level", x = "pe", color = color_column, label = "pe", group = "covs_dummy")) +
    geom_vline(xintercept = 0, color = "dark grey", cex = 1, alpha = 0.8) +
    facet_wrap(~trait, nrow = 2, labeller = labeller(trait = new_labels_traits)) +
    scale_y_discrete(limits = rev(levels(dataset$level)), labels = new_labels_levels) +
    scale_color_manual(values = color_9,
                       limits = levels(dataset$level),
                       labels = new_labels_levels) +
    geom_point(aes(shape = covs_dummy), size = 2, position = position_dodge(.6)) +
    geom_errorbarh(aes(xmin = pe - pe_se, xmax = pe + pe_se), height = 0.25, position = position_dodge(0.6), width = 0.1, size = 0.5) +
    scale_shape_manual(values = c(1, 16), 
                       limits = rev(levels(dataset$covs_dummy)), 
                       labels = c("No", "Yes"), 
                       guide = guide_legend(reverse = FALSE)) +
    scale_x_continuous(limits = c(-10, 119), oob = scales::oob_keep, labels = label_number(accuracy = 1), breaks = c(0, 25, 50, 75, 100)) +
    xlab(expression(paste("% of phenotypic variance explained (b"^2*")"))) +
    ylab("Data level") +
    ggtitle("OREML: Variance Component Analysis without covariates") +
    labs(shape = "Covariates", color = "Microbiome Measure (if q-value<0.05)") +
    theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
          legend.text = element_text(size = 30)) +
    geom_text(aes(label = ifelse(covs_dummy == "no_covs", as.character(round(pe, digits = 0)), '')), hjust = 0.6, vjust = 1.5, size = 3, position = position_dodge(.5)) +
    theme_light()
}
  
```


```{r}
## Function for secondary analyses - investigating mats and mia indication

make_secondary_plot_covs<-function(dataset, color_column){
  ggplot(dataset, aes(y = level, x = pe, group=covs_dummy, color=color_column, label=pe)) +
  geom_vline(xintercept = 0, color = "dark grey", cex = 1, alpha = 0.8) +
  facet_grid(trait~sample, labeller = labeller(sample=new_labels_samples)) +
  scale_y_discrete(limits=rev(levels(dataset$level)), labels = new_labels_levels) +
  scale_color_manual(values=color_9,
                     limits=(levels(dataset$level)),
                     labels=new_labels_levels) +
  geom_point(aes(shape = covs_dummy), size = 2.5, position=position_dodge(.6)) +
  geom_errorbarh(aes(xmin = pe-(pe_se), xmax = pe+(pe_se)), height = 0.25, position=position_dodge(0.6), width=0.1) +
  scale_shape_manual(values = c(1, 16), 
                     limits=rev(levels(subsamples2_adj$covs_dummy)), 
                     labels=c("No", "Yes"), 
                     guide = guide_legend(reverse = FALSE)) +
  scale_x_continuous(limits = c(-10, 100),oob=scales::oob_keep, labels = label_number(accuracy=1), breaks = c(0, 25, 50, 75, 100)) + # keeps the parts of error bars which are within the margins
  xlab(expression(paste("% of phenotypic variance explained (b"^2*")"))) +
  ylab("Data level") +
  ggtitle("OREML: Subsamples") +
  labs(shape= "Covariates", color = "Microbiome Measure (if q-value<0.05)") +
  theme(plot.title = element_text(size=20, face="bold", hjust=0.5),
        legend.text=element_text(size = 12),
        axis.text.x = element_text(size = 1),
        strip.text = element_text(size = 8)) +
  geom_text(aes(label=ifelse(covs_dummy=="yes_covs",as.character(round(pe, digits = 0)),'')), hjust=0.6, vjust=1.5, size=2.5, position=position_dodge(.8)) +
  theme_light()
}

```


# Primary and Secondary {.tabset}

- 9 microbiome measures, 174 participants
- take only clr-transformed measures

## No covs

```{r}

# Extract data
full_sample_no_covs <- data %>% 
  filter(sample == "174_sample",
         covs == "no_covs",
         trait %in% c("mia_indication", "mats_indication", "ASD", "age", "sex", "bmi", "bristol_regroup", "dietPC1", "dietPC2", "dietPC3"))  %>%
  mutate(trait = factor(trait, levels = c("mia_indication", "mats_indication", "ASD", "age", "sex", "bmi", "bristol_regroup", "dietPC1", "dietPC2", "dietPC3")),
  level = factor(level, levels = c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common", "Microba_common")))

# Plot
make_plot_covs(full_sample_no_covs,"sig_color")

ggsave("Output/OSCA/mia_forest_plot_full_sample_no_covs_not_adj_plots.pdf", width = 14, height = 8, dpi=600)

```

## With Covs 

```{r}

# Extract data
full_sample <- data %>% 
  filter(sample == "174_sample",
         covs %in% c("no_covs","age", "sex", "age_sex", "age_sex_dietPC"),
         trait %in% c("mia_indication", "mats_indication", "ASD", "age", "sex", "bmi", "bristol_regroup", "dietPC1", "dietPC2", "dietPC3")) %>% 
  mutate(trait = factor(trait, levels = c("mia_indication", "mats_indication", "ASD", "age", "sex", "bmi", "bristol_regroup", "dietPC1", "dietPC2", "dietPC3")),
         level = factor(level, levels = c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common", "Microba_common")),
         covs= factor(covs, levels = c("no_covs","age", "sex", "age_sex", "age_sex_dietPC")))

```

```{r}

# This selects the top and bottom covs values 
bottom_rows <- full_sample %>%                  # extract only the highest and lowest levels of covs
  group_by(level, sample, trait) %>% 
  arrange(desc(covs)) %>% 
  slice_tail(n = 1)
full_sample<-full_sample %>% 
  group_by(level, sample, trait) %>%  # and include them in the dataframe
  arrange(covs) %>% 
  slice_tail(n = 1) %>%
  bind_rows(bottom_rows) 

duplicated_rows <- full_sample[duplicated(full_sample), ] # check for duplicates
print(duplicated_rows) 

```

```{r}
# Plot
make_plot_covs(full_sample,"sig_color")
#ggsave("Output/OSCA/mia_forest_plot_selected_measures_9_covs_not_adj.pdf", width = 10, height = 6, dpi=600)

```

## FDR-adjusted
- Calcualting FDR for all traits
- FDR-adjusted p-value represents the probability of obtaining a result as extreme or more extreme than the observed result, while still maintaining the expected FDR level across all tests performed.

```{r}

# Calculation

levels <- unique(full_sample$level) # Get all unique levels
covs_list <- unique(full_sample$covs_dummy)

# Create an empty list to store the results
results <- list()

# Loop over each level
for (lvl in levels) {
  for (c in covs_list) {
  # Subset the data for the current level and "no_covs"
  subset_data <- subset(full_sample, level == lvl & covs_dummy == c)
  
  # Calculate FDR adjusted p-values
  subset_data$FDR <- p.adjust(subset_data$p_value, method = "fdr", n = 10)
  
  # Store the subset data in a new data frame with level name as column
  new_data <- subset_data
  new_data$level <- lvl
  new_data$covs_dummy <- c
  
  # Store the results in the list with level name as index
  results[[paste0(lvl, "_", c)]] <- new_data
  }
}

# Combine the results into a single data frame
full_sample_adj <- do.call(rbind, results)
full_sample_adj<-as.data.frame(full_sample_adj)
# Order levels
full_sample_adj <- full_sample_adj %>%
   mutate(level = factor(level, levels = c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common", "Microba_common")),
          covs_dummy = factor(covs_dummy, levels=c("yes_covs", "no_covs"))) # have to turn level into a factor, otherwise it doesn't work

# Take colour adjusted
full_sample_adj$sig_color_adj <- ifelse(full_sample_adj$FDR < (0.05), 
                                     as.character(full_sample_adj$level), "black") # in order to color depending on both level and significance

write_xlsx(full_sample_adj, "C:/Users/uqsvasil/Documents/PhD/MIA/Output/mia_osca_10_adjusted.xlsx")
write_xlsx(data, "C:/Users/uqsvasil/Documents/PhD/MIA/Output/mia_osca_all.xlsx")

# Plot
make_plot_covs(full_sample_adj,"sig_color_adj")
#ggsave("Output/OSCA/mia_forest_plot_selected_measures_9_covs_adj.pdf", width = 10, height = 6, dpi=600)

```

# Sensitivity analyses
- mia and mats indication
- sensitivity analyses when excluding participants based on meds intake/siblings

## Original

```{r}
# Extract data
subsamples <- data %>% 
  filter(covs %in% c("no_covs","age_sex_dietPC"),
         trait %in% c("mia_indication", "mats_indication")) %>% 
  mutate(trait = factor(trait, levels = c("mia_indication", "mats_indication")),
         covs_dummy = factor(covs_dummy, levels=c("yes_covs", "no_covs")),
         level = factor(level, levels = c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common","Microba_common")),
         sample = factor(sample, levels = c("174_sample", "full_bristol", "no_sib", "no_current_ab", "no_probiotic", "no_aap", "no_gastro_meds")))

datatable(subsamples)

# Plot
make_secondary_plot_covs(subsamples, "sig_color")

#ggsave("Output/OSCA/mia_forest_plot_subsamples_non_adj.pdf", width = 10, height = 6, dpi=600)

```

## Expanded
- true negative control and age groups analyses added

```{r}
# Extract data
subsamples2 <- data %>% 
  filter(covs %in% c("no_covs","age_sex_dietPC"),
         trait %in% c("mia_indication", "mats_indication")) %>% 
  bind_rows(data2) %>%
  mutate(trait = factor(trait, levels = c("mia_indication", "mats_indication")),
         covs_dummy = factor(covs_dummy, levels=c("yes_covs", "no_covs")),
         level = factor(level, levels = c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common","Microba_common")),
         sample = factor(sample, levels = c("174_sample", "full_bristol", "no_sib", "no_current_ab", "no_probiotic", "no_aap", "no_gastro_meds", "early_age_group", "middle_age_group", "late_age_group", "mats_excluded", "mia_excluded","acute_negative")))

datatable(subsamples2)

```

## FDR adjustment

```{r}

levels <- unique(subsamples2$level) # Get all unique levels
covs_list <- unique(subsamples2$covs_dummy)

# Create an empty list to store the results
results <- list()

# Loop over each level
for (lvl in levels) {
  for (c in covs_list) {
  # Subset the data for the current level and "no_covs"
  subset_data <- subset(subsamples2, level == lvl & covs_dummy == c)
  
  # Calculate FDR adjusted p-values
  subset_data$FDR <- p.adjust(subset_data$p_value, method = "fdr", n = 21)
  
  # Store the subset data in a new data frame with level name as column
  new_data <- subset_data
  new_data$level <- lvl
  new_data$covs_dummy <- c
  
  # Store the results in the list with level name as index
  results[[paste0(lvl, "_", c)]] <- new_data
  }
}

# Combine the results into a single data frame
subsamples2_adj <- do.call(rbind, results)
subsamples2_adj<-as.data.frame(subsamples2_adj)
# Order levels
subsamples2_adj <- subsamples2_adj %>%
   mutate(level = factor(level, levels = c("species_common", "species_rm0_rare", "genus_common", "family_common", "EClevel3_common", "EClevel4_common", "MetaCycpathway_common", "TCDB_common", "Microba_common")),
          covs_dummy = factor(covs_dummy, levels=c("yes_covs", "no_covs"))) # have to turn level into a factor, otherwise it doesn't work

# Take colour adjusted
subsamples2_adj$sig_color_adj <- ifelse(subsamples2_adj$FDR < (0.05), 
                                     as.character(subsamples2_adj$level), "black") # in order to color depending on both level and significance

write_xlsx(subsamples2_adj, "C:/Users/uqsvasil/Documents/PhD/MIA/Output/mia_osca_sensitivity_adjusted.xlsx")

```

```{r}

# Plot
make_secondary_plot_covs(subsamples2_adj, "sig_color_adj")

#ggsave("Output/OSCA/mia_forest_plot_subsamples2_adj.pdf", width = 10, height = 6, dpi=600)

```
